evidence of adaptation-independent time-warp spike timing noise correlations in cortical networks

### 2020-08-12 manuscript.tex
Lunch!

% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
\usepackage[colorinlistoftodos,disable]{todonotes}

@@

\section{Introduction}
The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population spike count \cite{reyes2015high, lin2015nature} or trial-averaged firing rate \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high} codes. Previous studies have shown that spike count variability is shared across the population \cite{scholvinck2015cortical, lin2015nature}. As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, in vivo shared variability is likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.

In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. Firstly, an additive factor is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. A second multiplicative factor also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down modulation... Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.

Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale fluctuations. In this paper, long-timescale refer to large changes taking place over multiple trials that are caused by neural adaptation in response to repetitive stimulation or possibly due to changes in the level of anaesthesia. Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.

@@

Rather than spike count noise correlations, this study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding, which controls for longer timescale changes in excitability such as neural adaptation. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
Particularly, a multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).
% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
% which approximates an underlying stimulus-response distribution.

@@

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.

@@

It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies.

A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. The spike time noise correlations uncovered in this study offer a new opportunity to study population-wide modulation by low-dimensional latent states and controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

Correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1). A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels. These correlated clusters represent reliable regions of response distribution space that allow analysis of spike timing noise correlations possibly caused by neural adaptation, a continuous non-adaptative latent state or a combination of both. A large proportion of these noise correlations would have been missed without clustering.

@@

Interestingly, many of the correlated clusters had non-unitary correlation angles. 
Importantly, clusters remained correlated with non-unitary gradient after controlling for neural adaptation,
% using time series analysis, 
giving evidence of precise non-adaptive noise correlations. Such non-unitary spike time correlation angles cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients are indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.

This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.

@@

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding. The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

>>>>

\usepackage[nomarkers,notablist,nofiglist]{endfloat}
\usepackage[colorinlistoftodos,disable]{todonotes}

@@

\section{Introduction}

The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}. Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}. Single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}. In vivo shared variability is therefore likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.

In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}. Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.

Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes. In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation or changes in the level of anaesthesia. Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.

@@

This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. 
% which would introduce the possibility of state-dependent encoding and decoding (see below). 
The characterisation controls for longer timescale changes in excitability to estimate the short-timescale state-modulated representation of a stimulus at a given point in time. 
% and a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron response can be expressed as a point in space, where each dimension represents the value of a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).

% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
% which approximates an underlying stimulus-response distribution.

@@

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).

% Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.

% Under such approaches,

@@

It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the short-timescale fluctuating state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
Long-term changes in cortical state, such as adpatation, could lead to correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising short-timescale or `adaptation-independent' noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, whilst controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. 
A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. 
% The spike time noise correlations uncovered in this study 
% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

Correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1). A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. It was also shown, however, that first spike pairs within these isolated clusters were positively correlated above chance-levels. These correlated clusters represent reliable regions of response distribution space that allow analysis of short- and long-timescale spike timing noise correlations. A large proportion of these noise correlations would have been missed without clustering.

@@

Interestingly, many of the correlated clusters had non-unitary correlation angles. 
Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
% using time series analysis, 
indicative of short-timescale fluctuating state that affects different neurons differently.
% giving evidence of precise non-adaptive noise correlations
% Such non-unitary spike time correlation angles cannot be 
This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).

This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons at a point in time are `warped' from trial-to-trial by some short-timescale fluctuating continuous cortical state.
% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.

@@

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between cluster. The work of this paper 
% in extracting and characterising adaptation-independent correlated clusters provides 
a useful component step towards investigating this. 
% confirming or rejecting state-dependent spike-time coding. 
The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
